Variance reduction for MCMC methods via martingale representations

### 2020-01-23 aap/Revision/vr-mcmc-revision.tex
update

\begin{thm}
\label{th:exp_mom_ula}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$ 
we have
%\begin{eqnarray*}
%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
%\end{eqnarray*}
 %and
%\begin{eqnarray*}
%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
%\end{eqnarray*}
%Moreover, 
\begin{eqnarray*}
\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray*}
for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$
\end{thm}
\begin{proof}

 Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
 \begin{align*}
\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
\end{align*}
\sergey{This theorem is not yet written in the second paper}
with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
\begin{align*}
|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
\end{align*}
we obtain
\begin{align*}
\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
\end{align*}

\begin{align*}
\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
\end{align*}
\end{proof}

\bibliographystyle{plain}
\bibliography{refs-1}
\end{document}

>>>>

\begin{thm}
\label{th:exp_mom_ula}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$ 
we have
%\begin{eqnarray*}
%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
%\end{eqnarray*}
 %and
%\begin{eqnarray*}
%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
%\end{eqnarray*}
%Moreover, 
\begin{eqnarray}
\label{eq:exp_decrease_step}
\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray}
for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover, 
\begin{eqnarray}
\label{eq:var_decrease_step}
\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
\end{eqnarray}
\end{thm}
\begin{proof}

 Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
 \begin{align}
 \label{eq:exp_bounded_dif}
\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
\end{align}
\sergey{This theorem is not yet written in the second paper}
with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
\begin{align*}
|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
\end{align*}
we obtain
\begin{align*}
\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
\end{align*}

\begin{align*}
\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
\end{align*}
\end{proof}
which proves \ref{eq:exp_decrease_step}. Note that from \ref{eq:exp_bounded_dif} by Taylor expansion for small $s$ we obtain
\begin{align*}
1 + \frac{s^2}{2}\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} + O(s^3) \leq 1 + s^2 \kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 + O(s^4)
\end{align*}
Since it holds for arbitarily small $s$, the variance is bounded by
\begin{align*}
\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} \leq 2\kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 \leq 2\kappa \Gamma_{l+1,l+n}
\end{align*}
Now we need to divide both parts by $\Gamma^2_{l+1,l+n}$ to get \ref{eq:var_decrease_step}. The proof is completed.
\bibliographystyle{plain}
\bibliography{refs-1}
\end{document}
